A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-08 latex/main.tex

The 5 Type II events in Table~\ref{tab:pango_x_lineages} 
are recombination events closely associated with Pango X lineages, 
but do not meet the strict origination criteria for Type I.
Type III events are Pango origin nodes in the ARG that are 
derived from Type I or Type II recombination events. There is significant
nesting among the Pango X lineage origin nodes within the ARG,
with two major clusters of particular note. Although XQ is 
directly associated with a recombination event in the ARG, 
XAA, XAG, XAM, XR and XU are all closely derived, suggesting 
that they are more parsimoniously explained as sublineages of XQ
than de-novo recombinant lineages.
The second cluster is based around a putative undocumented
recombinant that is the ancestor of lineages 
XZ, XAC, XAD, XAE, and XAP, which we refer to as ``Xx''
(Figure~\ref{fig:subgraphs}B).

The final class of events (Type IV) in Table~\ref{tab:pango_x_lineages}
corresponds to nodes that are not associated with a recombination event.
These vary in their plausibility of being non-recombinant.
For example, XN (Figure~\ref{fig:subgraphs}C) is only one mutation
distant from a BA.2 sample and has little additional evidence
to suggest recombination [ref sup section].
XB, on the other hand,
has been thoroughly characterized \cite{Gutierrez2022} and
may have misclassified by sc2ts due to limitations in
handling deletions or matching against long branches (see Discussion).
See Document~\ref{sec:supplement_indepth_pango_x_lineages}
for extended analyses on all of the Pango X lineages above,
Figure~\ref{fig:pango_x_copying_patterns} for copying patterns 
associated with Table~\ref{tab:pango_x_lineages}, and
Document S2 for ARG visualizations.

@@

\newpage
\setcounter{secnumdepth}{2} % Print out appendix section numbers

@@

\subsubsection{XA}
(also illustrated in Figure 2a / 3d)
XA traces to a well supported recombination node, with no reversions.
Note that the “causal” sample ERR5308556
(which triggers the initial recombination event)
is identical to the recombination node, and
therefore lacks C8090T,
which is one of the XA consensus mutations
(identified as those shared by over 90\% of XA samples).
As mentioned elsewhere,
the sc2ts detection results in terms of
the Pango lineages of the parents and the breakpoint intervals
agree with the results reported by Jackson et al. (\ref{tab:jackson}).
Two of the Jackson et al. XA samples
are present in the ARG (ERR5308556 and ERR5414941);
these samples are specifically marked in Figure~\ref{fig:subgraphs},
which is a summary of this subgraph.

\subsubsection{XB}
https://github.com/jeromekelleher/sc2ts-paper/issues/332

@@

\subsubsection{XC}
The five XC samples descend from a well supported recombination node.
The two parents that are both identical to actual samples
(ERR4908740 on the left and DRR321274 on the right).
There is one reversion (G21987A) on the left side of the breakpoint,
but the site 21987 was masked during ARG inference
because it was identified to be experiencing
frequent mutation in a preliminary ARG.

\subsubsection{XE/XH}
https://github.com/jeromekelleher/sc2ts-paper/discussions/633

@@

\subsubsection{XF}
XF is a well-supported recombinant between Delta and Omicron BA.1,
with the breakpoint on the far left of the genome.
All XF samples are the unique descendants of
a well supported recombination node,
with both parent nodes (ERR6193241: AY.4 and ERR8042040: BA.1) and
the single child node (ERR8076129) being samples.

\subsubsection{XG}
The three XG samples descend from a reconstructed internal node,
which in turn descends from a well supported recombination node.
The left parent is a sample (SRR20179863), and
the right parent is a reconstructed internal node
assigned to Omicron BA.2,
which is supported by a sample (ERR9124067) that is identical to it.

\subsubsection{XJ}
All the XJ samples descend from a well supported recombination node.
There are 14 samples (including e.g. ERR8683190)
which sc2ts identified as descendants of the same recombination node
but are labeled BA.2 by pangolin (displayed on the left), and
which do not possess the XJ consensus mutations C22792T and C27945T.
These BA.2 samples could be worthy of further investigation.

\subsubsection{XL}
All the XL samples descend from a well supported recombination node,
with three child samples which are identical to it.
The left parent is a sample (ERR7989424), and
the right parent is a reconstructed internal node
assigned to BA.2.5 with a large number of descendant samples.

\subsubsection{XM/XAL}
https://github.com/jeromekelleher/sc2ts-paper/issues/284

@@

\subsubsection{XN/XAU}
(also illustrated in Figure 2c)
https://github.com/jeromekelleher/sc2ts-paper/issues/348
https://github.com/jeromekelleher/sc2ts-paper/issues/358

@@

\subsubsection{XP}
https://github.com/jeromekelleher/sc2ts-paper/issues/345

@@

\subsubsection{XQ/XR/XU/XAA/XAG/XAM}
\label{sec:supplement_indepth_pango_x_lineages_xq}
https://github.com/jeromekelleher/sc2ts-paper/issues/338

@@

\subsubsection{XS}
https://github.com/jeromekelleher/sc2ts-paper/issues/287

@@

\subsubsection{XW}
All the XW samples descend from a well supported recombination node.
Its left parent is a BA.1.1.15-labeled sample (ERR8069658), and
its right parent is a BA.2-labeled sample (ERR9086669).

\subsubsection{XY}
All the XY samples descend from a well supported recombination node,
through a shared reconstructed internal node.
Its left parent is a reconstructed internal node assigned to BA.1.1,
which is supported by a sample (ERR8627048) identical to it.
Its right parent is a BA.2-labeled sample (ERR9399983).
There are a pair of (orange) reversion mutations
within the XY clade at position 23948:
these were not used for inference
but remapped using parsimony during postprocessing, and
indicate that the topology within the XY clade
could potentially be improved
by alternative phylogenetic reconstruction methods.

\subsubsection{Xx: XZ/XAC/XAD/XAE/XAP}
https://github.com/jeromekelleher/sc2ts-paper/discussions/631

This nested set of Pango X lineages descends
from a single recombination node in the ARG, and
we suggest that these Pango X lineages
may be aliases of a single recombinant
which we tentatively refer to as "Xx".
Note that all were designated around the same time,
with their samples collected
roughly within a three-month period (Mar., 2022 to Jul., 2022).
Several samples labeled as BA.2 by pangolin also lie
within this recombinant clade,
including two that are identical
to the proposed originating recombinant (ERR8146303 and ERR8163061):
we suggest that researchers examining the emergence of
these Pango X lineages could direct efforts
into examination of these and closely related samples.
The detection of Xx illustrates
the ability of sc2ts to clarify relationships between
samples during the course of a pandemic,
informing designation of emerging recombinants.

@@

\subsubsection{XAF}
https://github.com/jeromekelleher/sc2ts-paper/discussions/638

@@

\subsubsection{XAJ}
https://github.com/jeromekelleher/sc2ts-paper/discussions/634

@@

\subsubsection{XAN/XAV}
https://github.com/jeromekelleher/sc2ts-paper/discussions/783
https://github.com/jeromekelleher/sc2ts-paper/discussions/785

@@

\subsubsection{XAS}
https://github.com/jeromekelleher/sc2ts-paper/discussions/784

@@

\subsubsection{XAZ}
https://github.com/jeromekelleher/sc2ts-paper/discussions/786

@@

\subsubsection{XBB}
https://github.com/jeromekelleher/sc2ts-paper/issues/336

@@

\subsubsection{XBB.1}
XBB.1 is characterized by an additional mutation in Spike, G22317T (G252V) and
is not designated as a recombinant of XBB in the Pango designation issue
(https://github.com/cov-lineages/pango-designation/issues/1088).
The ARG agrees with this and places the characteristic mutation
above the root of a single non-recombinant XBB.1 clade.
We include this non-recombinant lineage
in order to highlight some well supported subsequent recombination nodes
within XBB.1, such as the (pink) one leading to two XBB.1-labeled samples
at the bottom left of the subgraph which we refer to as XBB.x.
XBB.x is well supported: its left parent is a reconstructed internal node,
which is directly supported by a BQ.1-labeled sample, and
its right parent is a BB.1.5.18-labeled sample (ERR10792363).
This highlights the automatic ability of sc2ts
to infer sequential recombination events
(i.e. recombinants-of-recombinants),
even those involving few samples.
A few other recombination events occur within XBB.1,
but they either fail QC checks, or
(as the other one shown here, to the bottom right)
lead to a single sample so may be
e.g. caused by accidental mixing of samples during sequencing.

\subsubsection{XBD}
All the XBD samples descend from a well supported recombination node.
Its left and right parents are reconstructed internal nodes,
both of which are directly supported by samples
(SRR21382561 and ERR9825609, respectively)
that are identical to them.

\subsubsection{XBE}
https://github.com/jeromekelleher/sc2ts-paper/issues/351

@@

\subsubsection{XBF}
All the XBF samples descend from a well supported recombination node,
with three mutations on the left of the breakpoint.
Its left parent is a sample assigned to BA.5.2.1 (SRR20305553).
Its right parent is a reconstructed internal node
that is supported by a CJ.1-labeled sample (SRR21948068),
which differs by three mutations.

\subsubsection{XBG}
All the XBG samples descend from a well supported recombination node,
with two mutations, one on each side of the breakpoint.
Both its left and right parents are samples
(SRR19920084 and ERR9852351, respectively).

\subsubsection{XBH}
All the XBH samples descend from a well supported recombination node,
despite two reversions (T222C and T12789C, respectively),
which occur at excluded sites and
were remapped by parsimony during postprocessing.
If these two sites had been used during ARG inference,
the branch attached to BA.2.1 would have been attached to
the BA.2.5 shown on the top left instead,
leading to a more parsimonious local topology
with two mutations on the left branch. This would also remove the two
blue lines on the left and side of the copying pattern.

\subsubsection{XBQ/XBK/XBK.1}
https://github.com/jeromekelleher/sc2ts-paper/issues/349
https://github.com/jeromekelleher/sc2ts-paper/issues/350

XBK and XBQ are siblings, and neither appear to be recombinants in the ARG,
as they are closely related to the non-recombinant node that originates CJ.1.3. 
At the time of designating XBK
a hypothesis was made that a missing common ancestor could exist 
(https://github.com/cov-lineages/pango-designation/issues/1440\#issuecomment-1355264973),
making XBK non-recombinant.
Exactly such a missing ancestor
is found as the upper internal node labeled CJ.1 in the ARG,
constructed by sc2ts as part of its tree building steps.
We suggest that sc2ts can therefore play an epidemiological role
both in generating hypotheses involving recombination, and
quantifying mutational evidence to support them.

\subsubsection{XBM}
All the XBM samples descend from a well supported recombination node.
Its left parent is a reconstructed internal node
that is directly supported by a BA.2.76-labeled sample identical to it.
Its right parent is a BF.3-labeled sample.

\subsubsection{XBR}
This one XBR sample descends from a well supported recombination node.
Its left parent is a BN.3.1-labeled sample, and
its right parent is a reconstructed internal node
that is directly supported by a BQ.1.25-labeled sample identical to it.

>>>>

The 5 Type II events in Table~\ref{tab:pango_x_lineages} 
are recombination events closely associated with Pango X lineages, 
but do not meet the strict origination criteria for Type I.
Type III events are Pango origin nodes in the ARG that are 
derived from Type I or Type II recombination events. There is significant
nesting among the Pango X lineage origin nodes within the ARG,
with two major clusters of particular note. Although XQ is 
directly associated with a recombination event in the ARG, 
XAA, XAG, XAM, XR and XU are all closely derived, suggesting 
that they are more parsimoniously explained as sublineages of XQ
than de-novo recombinant lineages.
The second cluster is based around a putative undocumented
recombinant that is the ancestor of lineages 
XZ, XAC, XAD, XAE, and XAP, which we refer to as ``Xx''
(Figure~\ref{fig:subgraphs}B, Document \ref{sec:dd_xx}).

The final class of events (Type IV) in Table~\ref{tab:pango_x_lineages}
corresponds to nodes that are not associated with a recombination event.
These vary in their plausibility of being non-recombinant.
For example, XN (Figure~\ref{fig:subgraphs}C) is only one mutation
distant from a BA.2 sample and has little additional evidence
to suggest recombination (Document \ref{sec:dd_xn}).
XB, on the other hand,
has been thoroughly characterized \cite{Gutierrez2022} and
may have misclassified by sc2ts due to limitations in
handling deletions or matching against long branches (see Discussion).
See Document~\ref{sec:supplement_indepth_pango_x_lineages}
for extended analyses on all of the Pango X lineages above,
Figure~\ref{fig:pango_x_copying_patterns} for copying patterns 
associated with Table~\ref{tab:pango_x_lineages}, and
Document S2 for ARG visualizations.

@@

\newpage
\setcounter{secnumdepth}{3} % Print out appendix section numbers

@@

\subsubsection{XA}
% mneumonic: deep dive XA
\label{sec:dd_xa}
(also illustrated in Figure 2a / 3d)
XA traces to a well supported recombination node, with no reversions.
Note that the “causal” sample ERR5308556
(which triggers the initial recombination event)
is identical to the recombination node, and
therefore lacks C8090T,
which is one of the XA consensus mutations
(identified as those shared by over 90\% of XA samples).
As mentioned elsewhere,
the sc2ts detection results in terms of
the Pango lineages of the parents and the breakpoint intervals
agree with the results reported by Jackson et al. (\ref{tab:jackson}).
Two of the Jackson et al. XA samples
are present in the ARG (ERR5308556 and ERR5414941);
these samples are specifically marked in Figure~\ref{fig:subgraphs},
which is a summary of this subgraph.

\subsubsection{XB}
\label{sec:dd_xb}
https://github.com/jeromekelleher/sc2ts-paper/issues/332

@@

\subsubsection{XC}
\label{sec:dd_xc}
The five XC samples descend from a well supported recombination node.
The two parents that are both identical to actual samples
(ERR4908740 on the left and DRR321274 on the right).
There is one reversion (G21987A) on the left side of the breakpoint,
but the site 21987 was masked during ARG inference
because it was identified to be experiencing
frequent mutation in a preliminary ARG.

\subsubsection{XE/XH}
\label{sec:dd_xexh}
https://github.com/jeromekelleher/sc2ts-paper/discussions/633

@@

\subsubsection{XF}
\label{sec:dd_xf}
XF is a well-supported recombinant between Delta and Omicron BA.1,
with the breakpoint on the far left of the genome.
All XF samples are the unique descendants of
a well supported recombination node,
with both parent nodes (ERR6193241: AY.4 and ERR8042040: BA.1) and
the single child node (ERR8076129) being samples.

\subsubsection{XG}
\label{sec:dd_xg}
The three XG samples descend from a reconstructed internal node,
which in turn descends from a well supported recombination node.
The left parent is a sample (SRR20179863), and
the right parent is a reconstructed internal node
assigned to Omicron BA.2,
which is supported by a sample (ERR9124067) that is identical to it.

\subsubsection{XJ}
\label{sec:dd_xj}
All the XJ samples descend from a well supported recombination node.
There are 14 samples (including e.g. ERR8683190)
which sc2ts identified as descendants of the same recombination node
but are labeled BA.2 by pangolin (displayed on the left), and
which do not possess the XJ consensus mutations C22792T and C27945T.
These BA.2 samples could be worthy of further investigation.

\subsubsection{XL}
\label{sec:dd_xl}
All the XL samples descend from a well supported recombination node,
with three child samples which are identical to it.
The left parent is a sample (ERR7989424), and
the right parent is a reconstructed internal node
assigned to BA.2.5 with a large number of descendant samples.

\subsubsection{XM/XAL}
\label{sec:dd_xm}
https://github.com/jeromekelleher/sc2ts-paper/issues/284

@@

\subsubsection{XN/XAU}
\label{sec:dd_xn}

(also illustrated in Figure 2c)
https://github.com/jeromekelleher/sc2ts-paper/issues/348
https://github.com/jeromekelleher/sc2ts-paper/issues/358

@@

\subsubsection{XP}
\label{sec:dd_xp}

https://github.com/jeromekelleher/sc2ts-paper/issues/345

@@

\subsubsection{XQ/XR/XU/XAA/XAG/XAM}
\label{sec:dd_xq}

https://github.com/jeromekelleher/sc2ts-paper/issues/338

@@

\subsubsection{XS}
\label{sec:dd_xs}

https://github.com/jeromekelleher/sc2ts-paper/issues/287

@@

\subsubsection{XW}
\label{sec:dd_xw}

All the XW samples descend from a well supported recombination node.
Its left parent is a BA.1.1.15-labeled sample (ERR8069658), and
its right parent is a BA.2-labeled sample (ERR9086669).

\subsubsection{XY}
\label{sec:dd_xy}

All the XY samples descend from a well supported recombination node,
through a shared reconstructed internal node.
Its left parent is a reconstructed internal node assigned to BA.1.1,
which is supported by a sample (ERR8627048) identical to it.
Its right parent is a BA.2-labeled sample (ERR9399983).
There are a pair of (orange) reversion mutations
within the XY clade at position 23948:
these were not used for inference
but remapped using parsimony during postprocessing, and
indicate that the topology within the XY clade
could potentially be improved
by alternative phylogenetic reconstruction methods.

\subsubsection{Xx: XZ/XAC/XAD/XAE/XAP}
\label{sec:dd_xx}

https://github.com/jeromekelleher/sc2ts-paper/discussions/631

This nested set of Pango X lineages descends
from a single recombination node in the ARG, and
These Pango X lineages
may be aliases of a single recombinant
which we tentatively refer to as ``Xx''.
Note that all were designated around the same time,
with their samples collected
roughly within a three-month period (Mar., 2022 to Jul., 2022).
Several samples labeled as BA.2 by pangolin also lie
within this recombinant clade,
including two that are identical
to the proposed originating recombinant (ERR8146303 and ERR8163061).

@@

\subsubsection{XAF}
\label{sec:dd_xaf}

https://github.com/jeromekelleher/sc2ts-paper/discussions/638

@@

\subsubsection{XAJ}
\label{sec:dd_xaj}

https://github.com/jeromekelleher/sc2ts-paper/discussions/634

@@

\subsubsection{XAN/XAV}
\label{sec:dd_xan}

https://github.com/jeromekelleher/sc2ts-paper/discussions/783
https://github.com/jeromekelleher/sc2ts-paper/discussions/785

@@

\subsubsection{XAS}
\label{sec:dd_xas}

https://github.com/jeromekelleher/sc2ts-paper/discussions/784

@@

\subsubsection{XAZ}
\label{sec:dd_xaz}

https://github.com/jeromekelleher/sc2ts-paper/discussions/786

@@

\subsubsection{XBB}
\label{sec:dd_xbb}

https://github.com/jeromekelleher/sc2ts-paper/issues/336

@@

\subsubsection{XBB.1}
\label{sec:dd_xbb.1}

XBB.1 is characterized by an additional mutation in Spike, G22317T (G252V) and
is not designated as a recombinant of XBB in the Pango designation issue
(https://github.com/cov-lineages/pango-designation/issues/1088).
The ARG agrees with this and places the characteristic mutation
above the root of a single non-recombinant XBB.1 clade.
We include this non-recombinant lineage
in order to highlight some well supported subsequent recombination nodes
within XBB.1, such as the (pink) one leading to two XBB.1-labeled samples
at the bottom left of the subgraph which we refer to as XBB.x.
XBB.x is well supported: its left parent is a reconstructed internal node,
which is directly supported by a BQ.1-labeled sample, and
its right parent is a BB.1.5.18-labeled sample (ERR10792363).
This highlights the automatic ability of sc2ts
to infer sequential recombination events
(i.e. recombinants-of-recombinants),
even those involving few samples.
A few other recombination events occur within XBB.1,
but they either fail QC checks, or
(as the other one shown here, to the bottom right)
lead to a single sample so may be
e.g. caused by accidental mixing of samples during sequencing.

\subsubsection{XBD}
\label{sec:dd_xbd}

All the XBD samples descend from a well supported recombination node.
Its left and right parents are reconstructed internal nodes,
both of which are directly supported by samples
(SRR21382561 and ERR9825609, respectively)
that are identical to them.

\subsubsection{XBE}
\label{sec:dd_xbe}

https://github.com/jeromekelleher/sc2ts-paper/issues/351

@@

\subsubsection{XBF}
\label{sec:dd_xbf}

All the XBF samples descend from a well supported recombination node,
with three mutations on the left of the breakpoint.
Its left parent is a sample assigned to BA.5.2.1 (SRR20305553).
Its right parent is a reconstructed internal node
that is supported by a CJ.1-labeled sample (SRR21948068),
which differs by three mutations.

\subsubsection{XBG}
\label{sec:dd_xbg}

All the XBG samples descend from a well supported recombination node,
with two mutations, one on each side of the breakpoint.
Both its left and right parents are samples
(SRR19920084 and ERR9852351, respectively).

\subsubsection{XBH}
\label{sec:dd_xbh}
All the XBH samples descend from a well supported recombination node,
despite two reversions (T222C and T12789C, respectively),
which occur at excluded sites and
were remapped by parsimony during postprocessing.
If these two sites had been used during ARG inference,
the branch attached to BA.2.1 would have been attached to
the BA.2.5 shown on the top left instead,
leading to a more parsimonious local topology
with two mutations on the left branch. This would also remove the two
blue lines on the left and side of the copying pattern.

\subsubsection{XBQ/XBK/XBK.1}
\label{sec:dd_xbq}

https://github.com/jeromekelleher/sc2ts-paper/issues/349
https://github.com/jeromekelleher/sc2ts-paper/issues/350

XBK and XBQ are siblings, and neither appear to be recombinants in the ARG,
as they are closely related to the non-recombinant node that originates CJ.1.3. 
At the time of designating XBK
a hypothesis was made that a missing common ancestor could exist 
(https://github.com/cov-lineages/pango-designation/issues/1440\#issuecomment-1355264973),
making XBK non-recombinant.
Exactly such a missing ancestor
is found as the upper internal node labeled CJ.1 in the ARG,
constructed by sc2ts as part of its tree building steps.

\subsubsection{XBM}
\label{sec:dd_xbm}

All the XBM samples descend from a well supported recombination node.
Its left parent is a reconstructed internal node
that is directly supported by a BA.2.76-labeled sample identical to it.
Its right parent is a BF.3-labeled sample.

\subsubsection{XBR}
\label{sec:dd_xbr}
This one XBR sample descends from a well supported recombination node.
Its left parent is a BN.3.1-labeled sample, and
its right parent is a reconstructed internal node
that is directly supported by a BQ.1.25-labeled sample identical to it.
